A simulation-based assessment of the relation between Stone Age sites and relative sea-level change along the Norwegian Skagerrak coast

### 2022-01-27 analysis/paper/paper.Rmd
Minor edits

Following the end of the Weichselian and the final retreat of the Fennoscandian Ice Sheet [e.g. @hughes2016; @stroeven2016], the isostatic rebound has been so severe that most areas of Norway have been subject to a continuous relative sea-level regression, despite corresponding eustatic sea-level rise [e.g. @lambeck1998; @svendsen1987]. In other words, the RSL has been dropping throughout prehistory. As this process is the result of glacial loading, the rate of uplift is more severe towards the centre of the ice sheet. Thus, some areas on the outer coast have had a more stable RSL or been subject to marine transgression [e.g. @romundset2015; @svendsen1987]. These conditions are directly reflected in the archaeological record. In areas where the sea-level has been stable over longer periods of time, people have often reused coastal site locations multiple times and over long time-spans, creating a mix of settlement events that are difficult to disentangle [@reitan2009]. Transgression phases, on the other hand, can lead to complete destruction of the sites, or bury them in marine sediments, leading to a hiatus in the archaeological record for certain sub-phases in the impacted areas [@bjerck2008]. Comparatively, given a continuous and still ongoing shoreline regression from as high as c. 220 m above present sea-level in the inner Oslo fjord, any one location in south-eastern Norway has only been shore-bound within a relatively limited time-span, and the sites have not been impacted by any transgressions [@hafsten1957; @hafsten1983; @sørensen1979]. This makes the region especially useful for evaluating the assumption of a shore-bound settlement over a long and continuous time-span.

@@

The shore-bound settlement location of prehistoric hunter-fisher-gatherers in Norway is generally believed to follow both from the exploitation of aquatic resources and from movement and communication, which would have been efficient on waterways. The same logic has also been extended to the hinter- and inland regions, where sites are to be located along rivers and lakes. This is to take a dramatic turn at the transition to the Late Neolithic, around 2400 BCE, with the introduction of the Neolithic proper. associated involving a change agro-pastoralism, the farm,  means that the site locations are now more withdrawn from the shoreline [@solheim2018; @prescott2012]. That is not to say that waterways and aquatic resources were no longer exploited, but rather that these activities would not have been as tightly integrated with settlement and tool-production areas as in preceding periods. Earlier, at the transition to the Early Neolithic (c. 4000 BCE) pottery is introduced to the sites, and there are some possible indications of an initial, small-scale uptake of agriculture at some sites in the Oslo fjord region. However, this initial uptake appears to be relatively small-scale and is believed to be combined with a continued and predominantly hunter-gatherer life-way, followed by a de-Neolithisation in the Middle Neolithic. [@nielsen2021] has recently argued that while there, the initial uptake of in the Early Neolithic is combined with a more complex settlement pattern, and argues that the simple forager/farming dichotomy has underplayed the variation present in the Early and Middle Neolithic data. The empirical expectation for this development would thus be a more varied relation between sites and the shore-line.

@@

Three of the sites have been associated with farming activities, either in directly or in the form building structures. The first is Nordby 1, which has . Furthermore, the Middle Neolithic phase at Kvastad A2 and Late Neolithic phase at Nauen A are both related to farming. Both of these sites also have ^14^C-dates and lithic inventory associated with Mesolithic forager activities. These agricultural phases are thus treated seperately below.  Finally, @nielsen2021 has recently suggested that Early and Middle Neolithic features from the otherwise younger sites Bratberg and Lang could be related to early agricultural activity in the Oslo fjord region.

@@

```{r shoredate_sites, echo = FALSE, fig.cap = "The result of shoreline dating the analysed sites by mean elevation of the site polygon, accounting for distance to closest isobases and employing the exponential decay function identified in Figure 7 for the distance between sites and shoreline. The shoreline dates in gray are plotted against the oldest radiocarbon-informed dates for each site in black. The red line indicates the 95% probability range for the shoreline date.", out.width = "525px", fig.align="center"}

@@

>>>>

Following the end of the Weichselian and the final retreat of the Fennoscandian Ice Sheet [e.g. @hughes2016; @stroeven2016; see Figure \@ref(fig:deglaciation)], the isostatic rebound has been so severe that most areas of Norway have been subject to a continuous relative sea-level regression, despite corresponding eustatic sea-level rise [e.g. @lambeck1998; @svendsen1987]. In other words, the RSL has been dropping throughout prehistory. As this process is the result of glacial loading, the rate of uplift is more severe towards the centre of the ice sheet. Thus, some areas on the outer coast have had a more stable RSL or been subject to marine transgression [e.g. @romundset2015; @svendsen1987]. These conditions are directly reflected in the archaeological record. In areas where the sea-level has been stable over longer periods of time, people have often reused coastal site locations multiple times and over long time-spans, creating a mix of settlement events that are difficult to disentangle [@reitan2009]. Transgression phases, on the other hand, can lead to complete destruction of the sites, or bury them in marine sediments, leading to a hiatus in the archaeological record for certain sub-phases in the impacted areas [@bjerck2008]. Comparatively, given a continuous and still ongoing shoreline regression from as high as c. 220 m above present sea-level in the inner Oslo fjord, any one location in south-eastern Norway has only been shore-bound within a relatively limited time-span, and the sites have not been impacted by any transgressions [@glorstad2020; @hafsten1957; @hafsten1983; @sørensen1979]. This makes the region especially useful for evaluating the assumption of a shore-bound settlement over a long and continuous time-span.

@@

The shore-bound settlement location of prehistoric hunter-fisher-gatherers in Norway is generally believed to follow both from the exploitation of aquatic resources and from movement and communication, which would have been efficient on waterways. The same logic has also been extended to the hinter- and inland regions, where sites are to be located along rivers and lakes. This is to take a dramatic turn at the transition to the Late Neolithic, around 2400 BCE, with the introduction of the Neolithic proper [@solheim2018; @prescott2020]. associated involving a change agro-pastoralism, the farm,  means that the site locations are now more withdrawn from the shoreline [@bakka1971; @prescott2012]. That is not to say that waterways and aquatic resources were no longer exploited, but rather that these activities would not have been as tightly integrated with settlement and tool-production areas as in preceding periods. At an earlier stage, at the transition to the Early Neolithic (c. 3900 BCE), pottery is introduced to the sites, and there are some possible indications of an initial, small-scale uptake of agriculture at some sites in the Oslo fjord region. However, this initial uptake appears to be relatively small-scale and is believed to be combined with a continued and predominantly hunter-gatherer life-way, followed by a complete de-Neolithisation in the Middle Neolithic. [@nielsen2021] has recently argued that while there, the initial uptake of in the Early Neolithic is combined with a more complex settlement pattern, and argues that the simple forager/farming dichotomy has underplayed the variation present in the Early and Middle Neolithic data. The empirical expectation for this development would thus be a more varied relation between sites and the shore-line from around 3900 BCE, and a decisive shift around 2400 BCE.

@@

Three of the sites have been associated with farming activities, either in directly or in the form building structures. The first is Nordby 1, which has . Furthermore, the Middle Neolithic phase at Kvastad A2 and Late Neolithic phase at Nauen A are both related to farming. Both of these sites also have ^14^C-dates and lithic inventory associated with Mesolithic forager activities. Following from the expected deviance from the settlement patterns that characterise forager sites, these agricultural phases are thus treated seperately below. Finally, @nielsen2021 has recently suggested that Early and Middle Neolithic features from the otherwise younger sites Bratberg and Lang could be related to early agricultural activity in the Oslo fjord region. Due to the uncertain nature of this assertion these are omitted here.

@@

```{r shoredate1, eval = FALSE, echo = FALSE, fig.cap = "The result of shoreline dating the analysed sites by mean elevation of the site polygon, accounting for distance to closest isobase and employing the exponential decay function identified in Figure 7 for the distance between sites and shoreline. The shoreline dates in gray are plotted against the oldest radiocarbon-informed phase for each site in black. The red line indicates the 95% probability range for the shoreline date.", out.width = "525px", fig.align="center"}

@@

```{r shoredate, echo = FALSE, fig.cap = "The result of shoreline dating the analysed sites by mean elevation of the site polygon, accounting for distance to closest isobase and employing the exponential decay function identified in Figure 7 for the distance between sites and shoreline. The shoreline dates in gray are plotted against the oldest radiocarbon-informed phase for each site in black. The red line indicates the $95%$ probability range for the shoreline date.", out.width = "525px", fig.align="center"}
knitr::include_graphics(here("analysis/figures/shoredate.png"))
```

# Shoreline dating the pioneer sites from Brunlanes

The pre-Boreal sites from the Brunlanes-project are among the earliest known sites in Norway [@jaksland2014a; @jaksland2012; @jaksland2012a]. They have a distinct Early Mesolithic artefact inventory and are situated in a steep area of the landscape where it would be difficult to envision any use of the sites after the sea retreated any significant distance from their location. Not least following from the fact that very few pre-Boreal ^14^C-dates associated with anthropogenic activity have been achieved in Norway, the shoreline dating of these earliest sites are of fundamental importance for understanding the pioneer settlement and intial colonisation of the Scandinavian peninsula [e.g. @glorstad2016; @bang-andersen2012]. In the original publication of the sites, Jaksland [@jaksland2014] provides a thorough discussion of shoreline dating in general and as used for the dating of the Brunlanes sites specifically. The result of comparing his results against the ones achieved using the above-outlined approach are given in Figure. These are largely in agreement. The sites are located more or less on the isobase of the displacement curve for Larvik, and neglible adjustments were therefore needed. The small discrepancies between the results can instead follow from the fact that a slightly updated version of the displacement curve is applied here. Jaksland's dates are given a 200 and 50 year uncertainty range starting from what he gives as the earliest possible date. The argument for the uncertainty range of only 50 years is based on the location of the sites in such a constrained and steep area of the landscape that the difference in elevation between the sites establish their relative date and thus constrain the uncertainty range so that they don't overlap between sites located at different elevations. This information is not integrated here, but could justify reducing the upper bound of the uncertainty range, especially for Pauler 1 and Pauler 2. While the accuracy is of course ultimately dependent on the veracity of the geological reconstruction, the steepness of the curve does nonetheless provide very precise dates, and the agreement with Jakslands thorough consideration speaks to general applicability of the approach to shoreline dating outlined in this study. This example also highlights the fact that a consideration of the surrouding terrain can help in reducing the date-span provided by the method, if certain sea-levels can be excluded as unlikely for when a site was in use. This makes less of an impact in this setting, where the 95% probability range is already limited, but would considerably improve the precision of the method in cases where the displacement curve is flatter (cf. Figure \@ref(fig:shoredate)).
